---
title: "Spatiotemporal modelling of White Hake populations in the Southern Gulf of the St-Lawrence"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(INLA)
library(raster)
library(patchwork)
library(sf)
library(tidyverse)
# library(inlatools) # remotes::install_github("inbo/inlatools")
# devtools::install_github("gfalbery/ggregplot")
library(ggregplot)
library(animation)
```

## Prepare the data

This data is STDIF structure - a class for unstructured spatio-temporal data, with n spatial locations and times, where n observations are available, from the `spacetime` package.

The 0s actually reflect that there were no observations at that time - in other words, they are _not_ false absences, and should not be removed.

```{r load data, echo = FALSE}
explan <- readRDS("data/nov_25/explan.RDS")
number <- readRDS("data/nov_25/number.RDS")
gulf <- readRDS("data/gulf.RDS")

# STDIF = space, time, data frame, i = irregular grid
head(number@sp)
head(number@time) # it has a time zone structure!!
head(number@endTime)
```

## Data exploration

### White hake abundances 

#### Through time

```{r plot-abundances}
plot(number@endTime, number@data$`WHITE HAKE`, type = "l")
```

#### In space

```{r map-abundances}
plot(number@sp, pch = 19, cex = number@data$`WHITE HAKE`/300)
plot(gulf, add = TRUE)
```

### Explanatory variables

```{r map-explan}
# function to add coordinates to dataset in preparation for conversion to sf
add_coords <- function(stidf_object){
  # assign datasets to objects to simplify the code a bit
  df <- stidf_object@data 
  coords <- stidf_object@sp@coords

  # add coordinate columns to dataset
  df <- mutate(df,
           longitude = coords[,1],
           latitude = coords[,2],
           time = lubridate::date(stidf_object@endTime),
           year = lubridate::year(stidf_object@endTime)
           )
  return(df)
}

# convert datasets to sf objects

explan_sf <- add_coords(explan) %>%
  st_as_sf(x = ., coords = c("longitude", "latitude"))
st_crs(explan_sf) <- explan@sp@proj4string # set coordinate ref system

gulf_sf <- st_as_sf(gulf)
st_crs(gulf_sf) <- gulf@proj4string # set coordinate ref system

# function ot make ggplot map of explanatory variable
explan_map <- function(variable, legend_title){
  ggplot() +
  geom_sf(data = gulf_sf, fill = "white") +
  geom_sf(data = filter(explan_sf, 
                        # pick a sequence of years to check out
                        year %in% c(1971, 1981, 1992, 2001, 2011, 2019)),
          aes(col = get(variable))) +
  facet_wrap(~year) +
  scale_color_viridis_c(option = "viridis") +
  labs(color = legend_title) +
  theme_void()
}

# maps
p1 <- explan_map("bottom.temperature", "Bottom \ntemperature")
p2 <- explan_map("bottom.salinity", "Bottom \nsalinity")
p3 <- explan_map("depth", "Depth")
p4 <- explan_map("slope", "Slope")
# patchwork together!
(p1 + p2) / (p3 + p4)
```



## Step 1 - Building the meshes

This mesh size reflects the range of coordinates in space. Changing to a smaller value, like 35000, gives a more appropriate mesh.

```{r build-meshes}
#_____________________
## Reorganise the data
#_____________________
### Space
xy <- coordinates(explan)
colnames(xy) <- c("locx", "locy")

### Time
time <- as.numeric(explan@endTime)/60 # Number of minutes since Jan 1, 1970 
timeBasis <- time - 884844 # Time starting at 1

#______________
## Build meshes
#______________
### Temporal mesh
timeGr <- 11
timeSeq <- seq(min(timeBasis),
               max(timeBasis),
               length = timeGr)

meshTime <- inla.mesh.1d(timeSeq)

### Spatial mesh
maxEdge <- 35000
meshSpace <- inla.mesh.2d(boundary = gulf, 
                          max.edge=maxEdge * c(0.5,2),
                          cutoff=maxEdge/4,
                          offset = c(10000,20000))

# plot points under the mesh
par(mar = c(1,1,1,1))
plot(number@sp, pch = 19, cex = 0.2, col = "red")
plot(gulf, add = TRUE)
plot(meshSpace, main = "", asp = 1, add = TRUE)

# plot abundances under the mesh too
plot(number@sp, pch = 19, cex = number@data$`WHITE HAKE`/300, col = "red")
plot(gulf, add = TRUE)
plot(meshSpace, main = "", asp = 1, add = TRUE)
```
```{r edge-number}
# Number of edges in the mesh
meshSpace$n
```

How many estimations will be carried out?
```{r complexity-mesh}
meshSpace$n * meshTime$n
```

## Step 2 - Define the stochastic partial differential equation

```{r spde, echo = FALSE}
SPDE <- inla.spde2.pcmatern(mesh = meshSpace, # where we want to estimate
                            alpha = 2, # must be between 0 and 2. this makes it curved rather than too sharp
                            prior.range = c(1000, 0.5), # distance at which the prior levels off;
                            # at a range of 100m, things should level off
                            # 0.5 means this is 50% likely. this gives a weight to the priors.
                            prior.sigma = c(35, 0.5) # amount of variance where it plateaus (slide 36/41)
                            )
# basis to build spatial structure of the model

# To set prior.sigma
sd(number@data$`WHITE HAKE`)
```

## Step 3 - Priors and hyperparameters for temporal autocorrelation

```{r priors}
# Temporal autocorrelation prior
hSpec <- list(theta=list(prior='pccor1', param=c(0.2, 0.9))) 
# assuming not a lot of temporal autocorrelation (0.2) and we are fairly sure of that (0.9)

# Precision likelihood 
## (negative binomial or any distribution that has more than one parameter)
precPrior <- list(prior='pc.prec', param=c(1, 0.01))
# guessing it's 1 with a confidence of .01, which is super low.

# note ^ is not useful if using a Poisson.
```


## Step 4 - Index matrix

```{r index-matrix}
Field <- inla.spde.make.index("field", 
                              n.spde = SPDE$n.spde,
                              n.group = meshTime$n)
```


## Step 5 - A matrix

```{r a-matrix}
# For estimation
A <- inla.spde.make.A(meshSpace,
                      loc = coordinates(number@sp),
                      group = timeBasis,
                      group.mesh = meshTime)
```

## Step 6 - Organise the A matrix into a list

```{r a-matrix-list}
Alist <- as.list(rep(1,5)) # list with single values
Alist[[1]] <- A
```


## Step 7 - Organise the effects (spatial autocorrelation structure and explanatory variables)

Here, we're using variables about the bottom conditions, because White hake is a groundfish.
```{r org-effects}
effect <- list(Field,
               bTemp =  explan@data$bottom.temperature,
               bSal = explan@data$bottom.salinity,
               bDep = explan@data$depth,
               strat = explan@data$stratum)
```


## Step 8 - Build stack

```{r build-stack}
Stack <- inla.stack(data=list(whitehake = number@data$`WHITE HAKE`),
                    A = Alist,
                    effects = effect,
                    tag="basis")
```

## Step 9 - Building the model (Finally!)

## Model selection

```{r build-candidate-models, eval = FALSE}
# build candidate models --- only run with two time points. otherwise, eternal
form_0 <- whitehake ~ 0 +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_1 <- whitehake ~ 0 + bTemp +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_2 <- whitehake ~ 0 + bSal +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_3 <- whitehake ~ 0 + bDep +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_4 <- whitehake ~ 0 + bTemp + bSal +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_5 <- whitehake ~ 0 + bTemp + bDep +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_6 <- whitehake ~ 0 + bSal + bDep +
  f(field, model=SPDE, group = field.group, 
    control.group=list(model='ar1', hyper=hSpec)) 
form_7 <- whitehake ~ 0 + bTemp + bSal + bDep +
  f(field, model=SPDE, group = field.group, # how they're grouped
    control.group=list(model='ar1', hyper=hSpec)) # build with model parameters
forms_ls <- list(form_0, form_1, form_2, form_3, form_4, form_5, form_6, form_7)

# evaluate each model
model_ls <- vector("list", length = length(forms_ls))
for(i in 1:length(model_ls)){
  model_ls[[i]] <- inla(forms_ls[[i]],
                        data = inla.stack.data(Stack),
                        family="gaussian",
                        control.family = list(link="identity"),
                        control.predictor = list(A = inla.stack.A(Stack), 
                                                 compute = TRUE, 
                                                 link = 1),
                        control.compute = list(waic = TRUE, config = TRUE))
}

# extract waic to compare
waic_df <- data.frame(model = 1:7, waic = NA)
for(i in 1:nrow(waic_df)){
  waic_df[i,2] <- model_ls[[i]]$waic$waic
}
# find lowest waic (it's model #4)
waic_df[which(waic_df$waic == min(waic_df$waic)),]

#saveRDS(model, "outputs/whitehake_model.RDS")
```

## Build chosen model
```{r build-model, eval = TRUE, cache = TRUE}
form <- whitehake ~ 0 + bTemp + bSal +
  f(field, # temporal autocorr
    model=SPDE, # spatial autocorr structure
    group = field.group, # how they're grouped
    control.group=list(model='ar1', hyper=hSpec) # build with model parameters
  ) +
  f(
    strat,
    model = "iid",
    hyper = list(theta = list(prior = "gaussian",
                              param = c(0, 0.001)))
  )

model <- inla(form,
              data = inla.stack.data(Stack),
              family="nbinomial",
              control.family = list(link="log", hyper = list(theta=precPrior)),
              control.predictor = list(A = inla.stack.A(Stack), 
                                       compute = TRUE, 
                                       link = 1),
              control.compute = list(waic = TRUE,
                                     config = TRUE))
```

## Checking and interpreting the model

```{r output-model}
summary(model)
model$summary.hyperpar
```

```{r check-model}
Efxplot(list(model)) + theme_classic()
# from Coding Club tutorial: "NB: There are no P values in INLA. Importance or significance of variables can be deduced by examining the overlap of their 2.5% and 97.5% posterior estimates with zero."
```

```{r check-residuals-Zuurbook}
ExpY <- model$summary.fitted.values[1:nrow(number@data),"mean"]

#For the variance we also need the posterior mean value of θ, which is obtained via
phi1 <- model$summary.hyper[1, "mean"]

#The variance is then given by
VarY <- ExpY * (1 - ExpY) / (1 + phi1)

# And once we have the mean and the variance we can calculate Pearson residuals.
E1 <- (number@data$`WHITE HAKE` - ExpY) / sqrt(VarY)

# Apply model validation
# Figure 23.11
par(mfrow = c(1,1), mar = c(5,5,2,2), cex.lab = 1.5)
plot(x = model$summary.fitted.values$mean[1:nrow(number@data)], 
     y = E1,
     xlab = "Fitted values",
     ylab = "Pearson residuals",
     xlim = c(0, 1))
abline(h = 0, lty = 2)
# note: there's a value at 130...
```


#### Fixed effects

This is the output you would get from a normal linear model.
- mode: mode of the parameters distribution
- kld: ????

#### Random effects

Field group is the random effect we set earlier.

#### DF

Distribution for the parameter picked with mean, sd, and credible interval

#### Predictions through space and time

```{r predict-model}
# Dimension of the raster
stepsize <- 1000 # choose cell size: 1000m x 1000m
rangeX <- range(meshSpace$loc[,1])
rangeY <- range(meshSpace$loc[,2])
nxy <- round(c(diff(rangeX), 
               diff(rangeY)) / stepsize)

# Define basis of the map
mapBasis <- inla.mesh.projector(meshSpace,
                               xlim = rangeX,
                               ylim = rangeY,
                               crs = crs(gulf))


# Calculate prediction
mapMean <- vector("list", length = timeGr)
map.025 <- vector("list", length = timeGr)
map.975 <- vector("list", length = timeGr)

# for every time point
for(i in 1:timeGr){
  # Model prediction
  fitMean <- inla.mesh.project(mapBasis, 
                          model$summary.random$field$mean[1:SPDE$n.spde + (i - 1) * SPDE$n.spde])
  fit.025 <- inla.mesh.project(mapBasis, 
                               model$summary.random$field$`0.025quant`[1:SPDE$n.spde + (i - 1) * SPDE$n.spde])
  fit.975 <- inla.mesh.project(mapBasis, 
                               model$summary.random$field$`0.975quant`[1:SPDE$n.spde + (i - 1) * SPDE$n.spde])

  # Build maps with confidence intervals
  mapMean[[i]] <- raster(t(fitMean[,ncol(fitMean):1]),
                         xmn = min(mapBasis$x), xmx = max(mapBasis$x), 
                         ymn = min(mapBasis$y), ymx = max(mapBasis$y),
                         crs = crs(gulf))

  map.025[[i]] <- raster(t(fit.025[,ncol(fit.025):1]),
                         xmn = min(mapBasis$x), xmx = max(mapBasis$x), 
                         ymn = min(mapBasis$y), ymx = max(mapBasis$y),
                         crs = crs(gulf))
  
  map.975[[i]] <- raster(t(fit.975[,ncol(fit.975):1]),
                         xmn = min(mapBasis$x), xmx = max(mapBasis$x), 
                         ymn = min(mapBasis$y), ymx = max(mapBasis$y),
                         crs = crs(gulf))
}  

# Convert to a rasterStack (aligns the elements of a list)
rasterMean <- stack(mapMean)
raster.025 <- stack(map.025)
raster.975 <- stack(map.975)
  # to check this visually: 
  # image(mapMean[[1]])
  # plot(meshSpace, add = TRUE)

# reorganise
for(i in 1:5){
  values(rasterMean)[,i] <- as.vector(t(mapMean[[i]]))
  values(raster.025)[,i] <- as.vector(t(map.025[[i]]))
  values(raster.975)[,i] <- as.vector(t(map.975[[i]]))
}

# Mask the region to keep only the region of interest
rasterMeanMask <- mask(rasterMean, gulf)
raster.025Mask <- mask(raster.025, gulf)
raster.975Mask <- mask(raster.975, gulf)
# image(rasterMeanMask)
# plot(gulf, add = TRUE)
```

```{r map-predictions, fig.height=4}
# Map the results
par(mfrow = c(2,3), mar = c(1,1,3,1))

for(i in 1:timeGr){
zlimRange <- range(values(raster.025Mask[[i]]),
                   values(rasterMeanMask[[i]]),
                   values(raster.975Mask[[i]]), na.rm = TRUE)
}

# convert to data frames for ggplot
raster.025Mask_df <- as.data.frame(raster.025Mask, xy = TRUE, na.rm = TRUE) %>%
  pivot_longer(cols = 3:ncol(.)) %>%
  mutate(quant = "Lower (0.025)")

rasterMeanMask_df <- as.data.frame(rasterMeanMask, xy = TRUE, na.rm = TRUE) %>%
  pivot_longer(cols = 3:ncol(.)) %>%
  mutate(quant = "Mean")

raster.975Mask_df <- as.data.frame(raster.975Mask, xy = TRUE, na.rm = TRUE) %>%
  pivot_longer(cols = 3:ncol(.)) %>%
  mutate(quant = "Higher (0.975)")

# bind together
raster_df <- rbind(raster.025Mask_df, rasterMeanMask_df, raster.975Mask_df)
raster_df$quant <- factor(raster_df$quant, levels = c("Lower (0.025)", "Mean", "Higher (0.975)"))

# make this more automatized later (label with years) 
raster_df$name <- factor(raster_df$name, levels = c("layer.1",
                                                    "layer.2",
                                                    "layer.3",
                                                    "layer.4",
                                                    "layer.5",
                                                    "layer.6",
                                                    "layer.7",
                                                    "layer.8", 
                                                    "layer.9", 
                                                    "layer.10", 
                                                    "layer.11"))

p <- ggplot() +
    geom_raster(data = filter(raster_df, quant == "Mean"), 
                aes(x = x, y = y, fill = value)) + 
    labs(fill = "Abundance") +
    scale_fill_distiller(palette = "RdBu", 
                         limits = c(-max(abs(raster_df$value)), 
                                    max(abs(raster_df$value)))) +
    #facet_wrap(~quant) +
    theme_void() +
  gganimate::transition_states(name, transition_length=.6, ) +
  labs(title = 'Time group: {closest_state}')
p
# save!
# gganimate::anim_save("predictions.gif", animation = p, path = "figures/")
```

Each time step:
```{r plot all time steps, fig.height = 13}
p2 <- ggplot() +
    geom_raster(data = raster_df, 
                aes(x = x, y = y, fill = value)) + 
    labs(fill = "Abundance") +
    scale_fill_distiller(palette = "RdBu", 
                         limits = c(-max(abs(raster_df$value)), 
                                    max(abs(raster_df$value)))) +
    facet_grid(name~quant) +
    theme_void() 
p2
```

